Performing and evaluating graph-based clustering with rOpenScPCA

### 2024-11-12 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
final touchups to complete the first draft of this notebook, and left some TODO comments to come back to

This notebook will demonstrate use of `rOpenScPCA` with the `SCPCS000001` sample from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory.
Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.

@@

- Louvain algorithm
- Jaccard weighting
- 10 nearest neighbors
- A resolution parameter of 1

@@

- The `algorithm` can be one of:
  - `"louvain"`, `"walktrap"`, or `"leiden"`
- The `weighting` can be one of:
  - `"jaccard"`, `"rank"`, or `"number"`
- The nearest neighbors parameter can be customized with the `nn` argument
- The resolution parameter can be customized with the `resolution` argument
  - This parameter is only used by Louvain and Leiden algorithms
- If the Leiden algorithm is used, its default objective function parameter will be `"CPM"`, but you can also set  `objective_function = "modularity"` instead.
- You can provide additional parameters as a list to the `cluster_args` argument.
  - Please refer to the [`igraph` documentation](https://igraph.org/r/html/latest) to learn more about what additional parameters can be provided to each clustering algorithm.
  - Note that `cluster_args` only accepts single-length arguments (no vectors or lists).

@@

- `cell_id`: Unique cell identifiers, e.g. cell barcodes
- `cluster`: A factor column with the cluster identities
- There will be one column for each clustering parameter used

Default use of this function is shown below.

@@

This section demonstrates how to use several functions for evaluating cluster quality and reliability. 
All examples will use the SCE object for demonstration, but a Seurat object can be provided instead and will provide the same results.
Again, as needed, you may need to specify the `pc_name` argument if the principal components do not use the object type's default name (`"PCA"` for SCE and `"pca"` for Seurat).

@@

This function takes the following arguments:
- An SCE or Seurat object that contains principal components
- A data frame with, at least, columns representing unique cell ids and cluster assignments.
  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)

This function will return the inputted data frame with two additional columns:
- `silhouette_width`: The calculated silhouette width for the cell
- `other`: The closet cluster to the cell besides the cluster to which it belongs, as used in the silhouette width calculation

@@

Looking at the full distribution of silhouette width values, we see most are positive which is encouraging, but the values are mostly below 0.5, and the mean is around 0.2 which is somewhat low.
Silhouette width therefore suggests that, while these clusters do have some structure, they are not very tight.

@@

```{r plot silhouette violin}
ggplot(silhouette_results) +
  aes(x = cluster, y = silhouette_width) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  # show the mean in value
  stat_summary(color = "red")
```

@@

This function takes the following arguments:
- An SCE or Seurat object that contains principal components
- A data frame with, at least, columns representing unique cell ids and cluster assignments.
  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)

This function will return the inputted data frame with two additional columns:
- `purity`: The neighborhood purity for the cell
- `maximum`: The cluster with the highest proportion of observations neighboring the cell

@@

With this view, we see that clusters 1, 3, and 6 all have average values below 0.75, but the rest are higher, in particular clusters 11, 12, and 13.
It's worth noting that both silhouette width and neighborhood purity found that clusters 11 and 12 were among the tightest clusters.

@@

- A value of 1 indicates they are completely overlapping
- A value of -1 indicates they are completely distinct
- A value of 0 indicates a random relationship

@@

We'll use the function `rOpenScPCA::calculate_stability` to calculate the stability of clusters using a bootstrapping approach, and then visualize the results.

This function takes the following arguments:
- An SCE or Seurat object that contains principal components
- A data frame with, at least, columns representing unique cell ids and cluster assignments.
  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)
TODO UPDATE HERE!!!!!!!!!!!!!!!!!!!!

- By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.
- This function will return a data frame with columns `replicate`, `ari`, and additional columns for the clustering parameters used when calculating bootstrap clusters.

Since this function uses random sampling when performing bootstrapping, it's important to pass in a random seed for reproducibility.

```{r stability, warning=FALSE}
stability_results <- calculate_stability(
  sce,
  cluster_results,
  seed = params$seed
)

@@

All bootstrap replicates produced values around 0.7, which is relatively close to 1. 
This result suggests that these clusters are fairly stable.

@@

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- calculate_stability(
  sce,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15,
  seed = params$seed
)

@@

Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv`), or you can add the results into your scRNA-seq object.

_A word of caution!_
Objects from the ScPCA Portal already contain a column called `cluster` with results from an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
To avoid ambiguity with these new clustering results, you may want to rename columns.

The subsequent examples will show you how to save the `cluster_results` data frame to a given object, as an example.
We'll begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results` data frame before adding them to the given object.

@@

Follow these steps save the clustering results to an SCE object's `colData`:

@@

```{r add to seurat}
# Add the cluster result data frame to the Seurat object metadata
seurat_obj <- AddMetaData(seurat_obj, cluster_results_openscpca)

>>>>

To demonstrate these tasks, this notebook will use the sample `SCPCS000001` from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory.
Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.

@@

* Louvain algorithm
* Jaccard weighting
* 10 nearest neighbors
* A resolution parameter of 1

@@

* The `algorithm` can be one of:
  * `"louvain"`, `"walktrap"`, or `"leiden"`
* The `weighting` can be one of:
  * `"jaccard"`, `"rank"`, or `"number"`
* The nearest neighbors parameter can be customized with the `nn` argument
* The resolution parameter can be customized with the `resolution` argument
  * This parameter is only used by Louvain and Leiden algorithms
* If the Leiden algorithm is used, its default objective function parameter will be `"CPM"`, but you can also set  `objective_function = "modularity"` instead.
* You can provide additional parameters as a list to the `cluster_args` argument.
  * Please refer to the [`igraph` documentation](https://igraph.org/r/html/latest) to learn more about what additional parameters can be provided to each clustering algorithm.
  * Note that `cluster_args` only accepts single-length arguments (no vectors or lists).

@@

* `cell_id`: Unique cell identifiers, e.g. cell barcodes
* `cluster`: A factor column with the cluster identities
* There will be one column for each clustering parameter used

@@

This section demonstrates how to use several functions for evaluating cluster quality and reliability. 
All examples will use the SCE object for demonstration, but a Seurat object can be provided instead.

All functions presented in this section take the following required arguments:

* An SCE or Seurat object that contains principal components
  * If the PCA matrix name is not the default for the object type (`"PCA"` for SCE and `"pca"` for Seurat), you can use `pc_name` to specify its name.
* A data frame with, at least, columns representing unique cell ids and cluster assignments.
  * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
  * You can override these defaults using the arguments `cell_id_col` and `cluster_col`

@@

This function will return the inputted data frame with two additional columns:

* `silhouette_width`: The calculated silhouette width for the cell
* `silhouette_other`: The closet cluster to the cell besides the cluster to which it belongs, as used in the silhouette width calculation

@@

Looking at the full distribution of silhouette width values, we see most are positive which is encouraging, but the values are mostly below 0.5, and the mean is less than 0.25.
Silhouette width therefore suggests that, while these clusters do have some structure, they are not very tight.

@@

```{r plot silhouette violin}
ggplot(silhouette_results) +
  aes(x = cluster, y = silhouette_width) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  # show the mean +/- SE in red
  stat_summary(color = "red")
```

@@

This function will return the inputted data frame with two additional columns:

* `purity`: The neighborhood purity for the cell
* `maximum_neighbor`: The cluster with the highest proportion of observations neighboring the cell

@@

In this view, we see that clusters 1, 3, and 6 all have average values below 0.75, but the rest are higher, in particular clusters 11, 12, and 13.
It's worth noting that both silhouette width and neighborhood purity found that clusters 11 and 12 were among the tightest clusters.

@@

* A value of 1 indicates they are completely overlapping
* A value of -1 indicates they are completely distinct
* A value of 0 indicates a random relationship

@@

We'll use the function `rOpenScPCA::calculate_stability` to calculate the stability of clusters using a bootstrapping approach, and then visualize the results.
By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.

This function will return a data frame with columns `replicate`, `ari`, and additional columns for the clustering parameters used when calculating bootstrap clusters.

Since this function uses random sampling when performing bootstrapping, it's also important to pass in a random seed for reproducibility with the argument `seed`.

```{r stability, warning=FALSE}
stability_results <- calculate_stability(
  sce,
  cluster_results$cluster, # TODO!!!!! Update once the rOpenScPCA hash is updated in this branch
  seed = params$seed
)

@@

All bootstrap replicates produced values around 0.7, which is relatively high and close to 1, which suggests that the these clusters are stable and robust to starting conditions.

@@

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- calculate_stability(
  sce,
  cluster_df_leiden$cluster, # TODO!!!!! Update once the rOpenScPCA hash is updated in this branch
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15,
  seed = params$seed
)

@@

Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv`), or you can add the results into your scRNA-seq object.
The subsequent examples will show you, as an example, how to save the `cluster_results` data frame to a given object.

_A word of caution!_
Objects from the ScPCA Portal already contain a column called `cluster` with results from an automated clustering.
These automatic clusters were not evaluated, and their parameters were not optimized for any given library.

To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.
We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results` data frame before adding them to a given object.

@@

Follow these steps to save the clustering results to an SCE object's `colData`:

@@

```{r add to seurat}
# Add the cluster result data frame to the Seurat object metadata
seurat_obj <- AddMetaData(seurat_obj, cluster_results_renamed)
